---
title: "PCA"
date: today
embed-resources: true
execute: 
  echo: false
  warning: false
  message: false   
format: html
---

```{r}
library(tidyverse)
library(DESeq2)
library(ggrepel)


all_mice <- readRDS("data/20251020-steve_expressionSet.rds")

```

```{r}
# Create DESeq2 dataset
dds0 <- DESeqDataSet(all_mice,
                        design = ~ 1 )
# Filter out genes with low expression (at least 10 counts in 5 or more samples)
keep <- rowSums(counts(dds0) >= 10) >= 5
dds <- dds0[keep,]
```

## PCA switch mice

Principal component analysis (PCA) of variance-stabilized RNA-seq counts from the switch cohort revealed that **tissue type** was the dominant driver of transcriptional variance. The first principal component (PC1), accounting for **98% of the total variance**, clearly separated **liver** and **brain** samples, indicating strong tissue-specific expression profiles. The second component (PC2), explaining about **1% of the variance**, captured minor within-tissue variability associated primarily with **sex**, with male and female samples forming partially distinct subclusters.

No obvious outliers were detected, and samples from independent batches overlapped within each tissue cluster, suggesting **minimal batch-related effects**. Together, these data confirm that tissue identity overwhelmingly dominates the transcriptional landscape in the switch mice.

### All mice

```{r}
#| fig-width: 6

# Principal component analysis (PCA) based on variance-stabilized counts
vsd_switch <- vst(dds[,dds$mice == "switch"], blind = FALSE)
pca_switch <- plotPCA(vsd_switch, intgroup = c("Sex", "group", "Tissue","batch","repeated"), returnData = TRUE)
percentVar_switch <- round(100 * attr(pca_switch, "percentVar"))

# PCA plot with aesthetics
ggplot(pca_switch, aes(PC1, PC2, color = Sex, shape = batch)) +
  geom_point(alpha = 0.6, size= 3) +
  xlab(paste0("PC1: ", percentVar_switch[1], "% variance")) +
  scale_shape_manual(values=c(1,8)) +
  ylab(paste0("PC2: ", percentVar_switch[2], "% variance")) +
  theme_minimal()
```

### PCA by tissue and sex

Within each tissue–sex subgroup, **Cre-AAV** and **GFP control** mice formed partially overlapping clusters, suggesting that the switch introduced **subtle transcriptomic changes relative to the global tissue effect**.

Importantly, samples from **independent experimental batches** overlapped extensively, indicating that batch effects were minimal after normalization.

```{r}
#| fig-height: 7
#| fig-width: 8
ggplot(droplevels(pca_switch), aes(PC1, PC2, color = group.1, shape = batch, label = repeated)) +
  geom_point( size= 3) +
  geom_text_repel(na.rm =T) +
  scale_shape_manual(values=c(1,8)) +
  theme_minimal()+
  facet_wrap(~ Tissue+Sex, scales = "free",ncol=2)
```

## PCA TR mice

### All TR mice

As occurred with switch mice, PCA of variance-stabilized RNA-seq counts from ApoE targeted-replacement (TR) mice revealed a strong separation of samples by **tissue type**, with **liver** and **brain** samples clearly segregating along the first principal component (PC1), which accounted for **98% of the total variance** .

```{r}
#| fig-width: 6
# Principal component analysis (PCA) based on variance-stabilized counts
vsd_TR <- vst(dds[,dds$mice == "TR"], blind = FALSE)
pca_TR <- plotPCA(vsd_TR, intgroup = c("Sex", "group", "Tissue","batch","repeated"), returnData = TRUE)
percentVar_TR <- round(100 * attr(pca_TR, "percentVar"))

# PCA plot with aesthetics
ggplot(pca_TR, aes(PC1, PC2, color = group.1, shape = Tissue)) +
  geom_point(alpha = 0.6, size= 3) +
  xlab(paste0("PC1: ", percentVar_switch[1], "% variance")) +
  scale_shape_manual(values=c(1,8)) +
  ylab(paste0("PC2: ", percentVar_switch[2], "% variance")) +
  theme_minimal()
```

### PCA by tissue with possible outliers labeled

Within the liver cluster, separation along the second principal component (PC2) distinguished **ApoE2/E2** mice from the **ApoE3/E3 and ApoE4/E4** groups . These data suggest that **ApoE2 expression has a more pronounced impact on liver transcriptomes** compared with the other ApoE isoforms, while tissue identity (liver vs. brain) remains the dominant source of variance across the dataset.

```{r}
#|fig-width: 8
#|fig-height: 2

ggplot(pca_TR |> 
    mutate(out = ifelse(PC2< (-15) | PC2>15 ,name,NA)),
     aes(PC1, PC2, color = group.1, label = out)) +
  geom_point(alpha = 0.6, size= 3) +
  xlab("PC1") +
  scale_shape_manual(values=c(1,8)) +
  ylab("PC2") +
  geom_text_repel(na.rm =T, size =3)+
  theme_minimal()+
  facet_wrap(~ Tissue, scales = "free")
```