---
title: "Gene Enrichment Analysis II"
subtitle: "Removed Mitochondria-Related Genes from DESeq2 Results"
date: today
embed-resources: true
execute: 
  echo: false
  warning: false
  messages: false
format: html
---


```{r}
# gprofiler2: Interface to g:Profiler for functional enrichment analysis
# https://biit.cs.ut.ee/gprofiler/gost
# https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
library(gprofiler2)

# fgsea: Fast Gene Set Enrichment Analysis
# https://bioconductor.org/packages/release/bioc/vignettes/fgsea/inst/doc/fgsea-tutorial.html
library(fgsea)

# msigdbr: Access to MSigDB gene sets for multiple species
library(msigdbr)

# Data manipulation and visualization
library(tidyverse)
library(kableExtra)
library(UpSetR)
```

To identify biological pathways and processes affected by Cre-mediated ApoE4→ApoE2 conversion in liver tissue, we performed complementary enrichment analyses using the high-confidence differentially expressed genes from Method 1 (combined batches, no batch adjustment).

```{r}

# Load complete differential expression results (all genes, not just significant)
liver_all <- read_csv("data/results_liver_1.csv")
liver <- liver_all |> 
  filter(str_starts(gene,"mt-|Mt-",negate = T))  # Remove mitochondrial genes


mitochondrial_genes <- setdiff(liver_all$gene, liver$gene)

```

_**The mouse mitochondrial genome encodes 37 genes in total:**_

* Protein-coding genes (13): ND1, ND2, COX1, COX2, ATP8, ATP6, COX3, ND3, ND4L, ND4, ND5, ND6, CYTB*

* Ribosomal RNA genes (2): 12S rRNA (RNR1), 16S rRNA (RNR2)*

* Transfer RNA genes (22): *tRNA-Phe (F), tRNA-Val (V), tRNA-Leu(UUR) (L1), tRNA-Ile (I), tRNA-Gln (Q), tRNA-Met (M), tRNA-Trp (W), tRNA-Ala (A), tRNA-Asn (N), tRNA-Cys (C), tRNA-Tyr (Y), tRNA-Ser(UCN) (S1), tRNA-Asp (D), tRNA-Lys (K), tRNA-Gly (G), tRNA-Arg (R), tRNA-His (H), tRNA-Ser(AGY) (S2), tRNA-Leu(CUN) (L2), tRNA-Thr (T), tRNA-Pro (P), tRNA-Glu (E)*

_**mitochondrial genes found (and removed) from the dataset:**_



```{r}
paste(mitochondrial_genes, collapse = ", ")
```


### Gene Set Enrichment Analysis (GSEA)


```{r}
#| fig-width: 10


#################################################################
### Mouse Gene Sets from MSigDB #################################
### https://www.gsea-msigdb.org/gsea/msigdb/mouse/genesets.jsp ##
#################################################################


  
# Create ranked gene list based on log2 fold change
# GSEA uses the ranking of ALL genes, not just significant ones
gene_ranks <- liver$log2FoldChange
names(gene_ranks) <- liver$gene
gene_ranks <- sort(gene_ranks, decreasing = TRUE)  # Sort high to low


set.seed(42)  # Set seed for reproducibility


# Retrieve Gene Ontology Biological Process gene sets from MSigDB
# MSigDB Collection M5 = ontology gene sets
# Subcollection BP = Biological Process
gene_sets <- msigdbr(
  species = "Mus musculus",      # Mouse
  db_species = "MM",              # Mouse species code
  collection = "M5",              # GO gene sets
  subcollection = "BP"            # Biological Process ontology
)

# Convert to list format required by fgsea
# Each element = one gene set (pathway) with its member genes
gene_sets_list <- split(gene_sets$gene_symbol, gene_sets$gs_name)

# Run Fast Gene Set Enrichment Analysis
# Tests whether genes in each pathway are randomly distributed in the ranked list
# or concentrated at the top (upregulated) or bottom (downregulated)
fgsea_results <- fgsea(
  pathways = gene_sets_list,      # Gene sets to test
  stats = gene_ranks,             # Ranked gene list
  eps = 0.0,                      # Precision for p-value estimation (0 = exact)
  minSize = 15,                   # Minimum genes per pathway
  maxSize = 500                   # Maximum genes per pathway
)

# Filter and format results
gsea <- fgsea_results |> 
  arrange(padj) |>                      # Sort by adjusted p-value
  filter(padj < 0.05) |>                # Keep only significant pathways
  select(-leadingEdge) |>                # Remove gene list column for display
  arrange(desc(NES))                     # Sort by normalized enrichment score
# Display results table
kbl(gsea |> select(pathway, NES, padj, size),
  digits = 3) |>
  kable_styling(full_width = FALSE, "striped")
```

Gene Set Enrichment Analysis (GSEA) using the complete ranked gene list (without the mitochondrial genes) identified `r nrow(gsea)` significantly enriched pathways. 

These pathways encompassed translation at the synapse (NES = 2.092, padj = 0.031), mitochondrial respiratory chain complex assembly (NES = 2.024, padj = 0.026), and NADH dehydrogenase complex assembly (NES = 1.990, padj = 0.037), suggesting that hepatic ApoE2 expression modulates both synaptic protein synthesis and mitochondrial oxidative capacity.

## Brain
### Gene Set Enrichment Analysis (GSEA)



```{r}
#| fig-width: 10
#| fig-height: 7


#################################################################
### Mouse Gene Sets from MSigDB #################################
### https://www.gsea-msigdb.org/gsea/msigdb/mouse/genesets.jsp ##
#################################################################

set.seed(42)  # Set seed for reproducibility

# Load complete differential expression results (all genes, not just significant)
brain <- read_csv("data/results_brain_1.csv")|> 
  filter(str_starts(gene,"mt-|Mt-", negate = T))  # Remove mitochondrial genes

# Create ranked gene list based on log2 fold change
# GSEA uses the ranking of ALL genes, not just significant ones
gene_ranks <- brain$log2FoldChange
names(gene_ranks) <- brain$gene
gene_ranks <- sort(gene_ranks, decreasing = TRUE)  # Sort high to low

# Retrieve Gene Ontology Biological Process gene sets from MSigDB
# MSigDB Collection M5 = ontology gene sets
# Subcollection BP = Biological Process
gene_sets <- msigdbr(
  species = "Mus musculus",      # Mouse
  db_species = "MM",              # Mouse species code
  collection = "M5",              # GO gene sets
  subcollection = "BP"            # Biological Process ontology
)

# Convert to list format required by fgsea
# Each element = one gene set (pathway) with its member genes
gene_sets_list <- split(gene_sets$gene_symbol, gene_sets$gs_name)

# Run Fast Gene Set Enrichment Analysis
# Tests whether genes in each pathway are randomly distributed in the ranked list
# or concentrated at the top (upregulated) or bottom (downregulated)
fgsea_results <- fgsea(
  pathways = gene_sets_list,      # Gene sets to test
  stats = gene_ranks,             # Ranked gene list
  eps = 0.0,                      # Precision for p-value estimation (0 = exact)
  minSize = 15,                   # Minimum genes per pathway
  maxSize = 500                   # Maximum genes per pathway
)



# Filter and format results
gsea <- fgsea_results |> 
  arrange(padj) |>                      # Sort by adjusted p-value
  filter(padj < 0.05) |>                # Keep only significant pathways
  arrange(desc(NES))                     # Sort by normalized enrichment score



# Display results table
kbl(gsea |> select(pathway, NES, padj, size),
  digits = 3) |>
  kable_styling(full_width = FALSE, "striped")
```

GSEA of brain tissue revealed `r nrow(gsea)` significantly enriched pathways.

#### Reducing Redundant Pathways with `collapsePathways`


Given the extensive overlap in leading-edge genes between pathways, we applied the `collapsePathways` function from the `fgsea` package to reduce redundancy and identify independent biological themes. This analysis groups pathways with highly similar gene membership, retaining only the most significantly enriched representative from each cluster.


```{r}

# Subset gene_sets_list to only include significant pathways
gene_sets_significant <- gene_sets_list[gsea$pathway]


# Step 3: Run collapsePathways
# This identifies redundant pathways based on leading edge overlap
collapsed <- collapsePathways(
  fgseaRes = fgsea_results |> filter(padj<0.05),     # Your fgsea results (filtered)
  pathways = gene_sets_significant,     # Gene sets (filtered to significant)
  pval.threshold = 0.01,               # P-value threshold for significance
  stats = gene_ranks                    # Same ranked gene list used in fgsea
)

# Step 4: Extract the main (representative) pathways
main_pathways <- collapsed$mainPathways

# Step 5: Filter your results to keep only main pathways
gsea_collapsed <- fgsea_results |> 
  filter(padj<0.05) |>
  filter(pathway %in% main_pathways) |>
  arrange(desc(NES))


# Display results table
kbl(gsea_collapsed |> select(pathway, NES, padj, size),
  digits = 3) |>
  kable_styling(full_width = FALSE, "striped")

```

This analysis yielded `r nrow(gsea_collapsed)` representative pathways that captured the major biological processes altered by hepatic ApoE2 expression. The collapsed pathway set retained the most significantly enriched representatives, including chromatin organization and remodeling, xenobiotic and metabolic processes, splicing and mRNA processing, neuronal development, and synaptic function, thereby providing a streamlined view of the fundamental biological processes affected by selective hepatic ApoE isoform conversion.